---
title: "Assignement 2"
output: "pdf_document"
date: "2024-01-10"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Introduction :

This assignment consists of performing data cleaning and manipulation, and then some statistical analysis. 

## Dataset:

The dataset is retrieved from [Rwanda DHS (Demographic and Health Survey)](https://dhsprogram.com/Data/) 2020. The type of dataset used here is Household member. You will get data in two files: main SPSS File and Map File (for descriptions). 

Your Assignments steps:

1. Read the dataset in R.

```{r}
options(warn = -1)
# required package
library(haven)

# import data
houseH <- read_sav("DHS2020/RWPR81FL.sav")


# data dimension
dims <-dim(houseH)


```
The dataset has `r dims[1]` observations and `r dims[2]` features

- Visualize, inspect and get familiar with the data
```{r}
# your code here

# top 6 rows
head(houseH)

# last 6 rows
tail(houseH)

# column names
column_names <- names(houseH)

statistical_summary <- summary(houseH)  # summary statistics for each column

#str(houseH) # Understanding the data structure (types, dimensions, column details).



```



2. Select only few columns, important in this Assignments. They are the following: "HV001", "HV009", "HV010", "HV011", "HV014", "SHDISTRICT", "HV024", "HV025", "HV040", "HV227", "HV228", "HV270", "HV105", "HV106", "HML3", "HML4", "HML7",
  "HML10", "HML22", "HML32","HML33", "HML35"

```{r}
# your code here
columns <- c("HV001", "HV009", "HV010", "HV011", "HV014", "SHDISTRICT", "HV024", "HV025", "HV040", "HV227", "HV228", "HV270", "HV105", "HV106", "HML3", "HML4", "HML7",
  "HML10", "HML22", "HML32","HML33", "HML35")
df <- houseH[,columns]
```

  
3. Rename variables using the variable descriptions below. Give meaningful (short) name to the variables of your choice.

>* HV001= "Cluster number", -> "cluster_no"
>* HV009 = "Number of household members", -> "no_HH_member"
>* HV010 = "Number of eligible women in household", -> "no_elig_wom_HH"
>* HV011 = "Number of eligible men in household", -> "no_elig_men_HH"
>* HV014 = "Number of children 5 and under (de jure)", -> "child_under_5years"
>* SHDISTRICT = "District (geographic area)", -> "district"
>* HV024 = "Region (provinces, corresponding values in a map file)", -> "province"
>* HV025 = "Type of place of residence (rural versus urban)", -> "locality
>* HV040 = "Cluster altitude in meters", -> "cluster_alt_M"
>* HV227 = "Presence of mosquito bed net for sleeping", -> "mosquito_net_yes"
>* HV228 = "Number of children under 5 who slept under a mosquito bed net", -> "child_under_5years_net_yes
>* HV270 = "Wealth index combined (an index based on various household assets indicating socio-economic  status)", -> "socio-economic_status
>* HV105 = "Age of household members", -> "HH_members_ages"
>* HV106 = "Highest educational level attained by individuals", -> "education_level"
>* HML3 = "Net observed by interviewer", -> "net_observed"
>* HML4 = "Months ago the net was obtained", -> "months_since_net_obtained"
>* HML7 = "Brand of net", -> "net_brand"
>* HML10 = "Insecticide-Treated Net (ITN)", -> "ITN"
>* HML22 = "Obtained net from campaign, antenatal, or immunization visit", -> 
>* HML33= "Result of malaria measurement", -> "malaria_measures"
>* HML32 = "Final result of malaria from blood smear test", -> "malaria_blood_T_result"
>* HML35 = "Result of malaria rapid test" -> "malaria_rapid_T_result"

```{r}
new_names <- c("cluster_no","no_HH_member","no_elig_wom_HH","no_elig_men_HH","child_under_5years","district",
               "province","locality","cluster_alt_M","mosquito_net_yes","child_under_5years_net_yes",
               "socio-economic_status","HH_members_ages","education_level","net_observed",
               "months_since_net_obtained","net_brand","ITN","antenatal","malaria_blood_T_result",
               "malaria_measures","malaria_rapid_T_result")


names(df) <- new_names

```




## Data cleaning

1. Inspect each variables, decode variable to its original unique variables. 
Example, Variable "HV024"(Region) has Unique values 1,2,3,4,5. Decode it to orginal Region Kigali, South, West, North, East
Use Map file to see the description of each values in data. 

```{r}

data.frame(sapply(df, class))

```
Upon examining the column classes, we discovered that some were labeled as `haven_labelled`. Referring to the **map file**, we chose to decode these columns to their original values using the `as_factor` method.

```{r}

decode <- function(data){
  for(name in colnames(data)){
    if("haven_labelled" %in% class(data[[name]]) & (!name %in% c("HH_members_ages","months_since_net_obtained"))){
    data[name] <- as_factor(data[name])
   }
  }
 return(data)
}
  
df <- decode(df)
```


2. Handling Missing Values:

Determine columns with missing values. Devise the strategy to handle missing values: Deleting missing values, replacing missing values with mean or mode.
```{r}
options(warn = -1)
# your code
library(dplyr)
count_NA <- function(data){
  data <- sapply(data, function(x) sum(is.na(x))) %>% sort(decreasing = TRUE) %>% data.frame() %>% rename( "null_count" = 1) %>% mutate(null_proprition=paste0(round(null_count/nrow(df)*100,2),"%"))
  
  return(data)
}
count_NA(df)
```
From the table above, it is evident that some columns have a significant proportion of missing values, ranging from **42%** to **80%**. Typically, columns with **80%** missing data would be discarded. However, as these columns are critical to our analysis, we opted to impute the missing values. For numerical columns, we used the **mean**, while for categorical columns, we used the **mode**.


```{r}
NA_imputer <- function(data){
  for(cname in colnames(Filter(function(col)  any(is.na(col)), data))){
    #print(paste(cname,":",names(sort(table(data[cname]), decreasing = T)[1])))
    if("factor" %in% class(data[[cname]])){
        data[cname][is.na(data[cname])] <- names(sort(table(data[[cname]]), decreasing = T)[1])

    }else{
      data[cname][is.na(data[cname])] <- as.integer(mean(data[[cname]], na.rm = T))

    }
  }
  return(data)
}

clean_df <- NA_imputer(df)

count_NA(clean_df)

```
After imputation, all columns are now free of `missing values`.



3. Create new variables

>a. Create variable called "Old Mosquito" variable HML4 (Months ago the net was obtained). The created variable must binary with 1 when mosquito is more than 24 months old. 
>b. Create Variable "Average District altitude". Create this variable by averaging cluster altitude in each district. We have three variables HV001= "Cluster number", SHDISTRICT = "District (geographic area)" and HV040 = "Cluster altitude in meters". Filter out clusters in each district, do `mean` of cluster altitude in that district. 

```{r}
library(dplyr)
clean_df <- clean_df %>%
  mutate(
    'Old Mosquito' = ifelse(months_since_net_obtained > 24, 1, 0)
  ) %>%
  group_by(district) %>%
  mutate(
    'Average District Altitude' = mean(cluster_alt_M, na.rm = TRUE)
  ) %>%
  ungroup()

```


## Data visualizations: 

Produce visualization of your choice. At least each of these
- Bar plot

```{r}
# your code
library(ggplot2)

clean_df %>% 
  group_by(province, mosquito_net_yes) %>% 
  summarise(total_nets = n(), .groups = "drop") %>% 
  ggplot(aes(x = province, y = total_nets, fill = as.factor(mosquito_net_yes))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Household with Mosquito Nets Distribution by Province", 
    x = "Province", 
    y = "Households", 
    fill = "Has Mosquito Net  Yes/No"
  ) +
  geom_text(
    aes(label = total_nets),
    position = position_dodge(width = 0.9),
    vjust = -0.5,
    size = 3
  ) +
  theme_minimal()


```
1.**Regional Distribution**

The **East Province** has the highest number of households with mosquito nets (`8,521`), followed closely by the ***South Province** (`9,289`).
The **North Province** has a relatively lower number of households with mosquito nets compared to other provinces.

2.**Urban vs Rural Trends**

In **Kigali**, a largely urban area, there is a stark contrast, with very few households (509) without mosquito nets. This could indicate better coverage programs or accessibility in urban settings.
```{r}
library(ggplot2)
clean_df %>% 
  group_by(province, education_level) %>% 
  summarise(ed_lev_counts = n(), .groups = "drop") %>% 
  ggplot(aes(x = province, y = ed_lev_counts, fill = as.factor(education_level))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Education Level Distribution by province", 
    x = "province", 
    y = "Count", 
    fill = "Education Levels"
  ) +
  geom_text(
    aes(label = ed_lev_counts),
    position = position_dodge(width = 0.9),
    vjust = -0.5,
    size = 2.5
  ) +
  theme_minimal()
```
The graph highlights that `primary education` dominates across all provinces, with the **South ** (8,072) and  **East** (7,564) provinces leading in numbers, reflecting a strong emphasis on basic education.  **Kigali ** stands out with a relatively higher number of individuals with `higher education` (619), showcasing its urban advantage, while other provinces (North, West, South, and East) lag significantly, with higher education counts below 300, indicating limited access to advanced education. The  **North ** (5,053) and  **East ** (3,836) provinces have a notable number of individuals with `no education` or `preschool-level` education. Lastly,  **Don't Know ** responses are minimal, suggesting respondents whom we consider as those did not want to give the information.
```{r}
clean_df %>% 
  group_by(locality, `socio-economic_status`) %>% 
  summarise(counts = n(), .groups = "drop") %>% 
  ggplot(aes(x = locality, y = counts, fill = as.factor(`socio-economic_status`))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Social Economic Status by Locality", 
    x = "Locality", 
    y = "Count", 
    fill = "Social Economic Status"
  ) +
  geom_text(
    aes(label = counts),
    position = position_dodge(width = 0.9),
    vjust = -0.5,
    size = 2.5
  ) +
  theme_minimal()
```
The **rural population** is predominantly concentrated in the `poorest (10,700)` and `poorer (10,440)` categories, highlighting widespread poverty in these areas. In contrast, **urban localities** have a significantly higher representation in the ``richest category (7,004)`. The `middle` and `richer` categories also show a stronger presence in **rural areas**, though the **urban population** contributes a smaller yet notable portion to these categories.

- Pie plot
```{r}
# your code
props <- round(prop.table(table(clean_df$locality)) * 100,2)
 props %>% pie(  main = "Distribution of Household by Locality",
labels =paste(names(props), "(",as.numeric(props),"%)")) 
 

```
The majority of households in survey are in rural areas.
`r as.numeric(props[2])`% of households are located in rural areas, while only `r as.numeric(props[1])`% are in urban areas.


```{r}

 #library(dplyr)

# Filter and calculate percentages for "Rural" locality only
#props1 <- clean_df %>%
 # filter(locality == "Rural") %>%
  #$count(antenatal) %>%
  #mutate(percentage = round((n / sum(n)) * 100, 2))

#props2 <- clean_df %>%
 # filter(locality == "Urban") %>%
  #count(antenatal) %>%
  #mutate(percentage = round((n / sum(n)) * 100, 2))

# Create pie chart
#windows(width = 12, height = 6)
#par(mfrow = c(1, 2))

# First pie chart
#pie(
 # props1$percentage,
 # main = "Rural Locality",
  #labels = paste(props1$antenatal, "(", props1$percentage, "%)", sep = ""),
  #col = rainbow(length(props1$percentage)), # Distinct colors for each slice
  #cex = 0.5, # Adjust label size
  #radius = .4
#)

# Second pie chart
#pie(
  #props2$percentage,
  #main = "Urban Locality",
  #labels = paste(props2$antenatal, "(", props2$percentage, "%)", sep = ""),
  #col = rainbow(length(props2$percentage)), # Distinct colors for each slice
  #cex = 0.5, # Adjust label size
  #radius = .4
#)

# Add a common title
#mtext("Comparison of Antenatal Care in Rural and Urban Localities", side = 3,line=3, cex = .95)

```


- Histogram
```{r}
# your code
hist(clean_df$HH_members_ages, main = "Distribution of Household Ages", xlab = "Ages")
```

- Boxplot
```{r}
# your code
boxplot(clean_df$HH_members_ages)
```


## Statistical analysis

### Descriptive statistics

1. Use Variable "HML33" to filter out people who had Malaria measurement. 
```{r}
# your code
table(clean_df$malaria_measures)['Measured']

```

2. Calculate Malaria [Prevalence](https://dhsprogram.com/data/Guide-to-DHS-Statistics/Prevalence_of_Malaria_in_Children.htm) for both "Blood Smear" and "Rapid Test"
```{r}
# your code
table(clean_df$malaria_rapid_T_result)
round(table(clean_df$malaria_blood_T_result)['Positive']/length(clean_df$malaria_blood_T_result) * 100, 2) 


round(table(clean_df$malaria_rapid_T_result)['Positive']/length(clean_df$malaria_rapid_T_result) * 100, 2) 

```

3. Aggregate Prevalence at district Level
```{r}
# your code
dd <- clean_df %>% group_by(district) %>% mutate(
  RPT_prev = round(table(malaria_rapid_T_result)['Positive']/length(malaria_rapid_T_result) * 100,2 ),
  BT_prev = round(table(malaria_blood_T_result)['Positive']/length(malaria_blood_T_result) * 100,2 )
                                                 
                                                 )



library(dplyr)

district_pre <- clean_df %>%
  group_by(district) %>%
  summarise(
    malaria_rapid_prevalence = round(sum(malaria_rapid_T_result == "Positive") / n() * 100, 2),
    malaria_blood_prevalence = round(sum(malaria_blood_T_result == "Positive") / n() * 100, 2)
  )


c <- c("district","RPT_prev","BT_prev")
unique(dd[c])

```

### Analytical Analysis
1. Compare the prevalence in both tests and state if they are different.

***Hint:*** Check `?` the documentations for `t.test` and `aov`.
```{r}
# your code
t.test(district_pre$malaria_rapid_prevalence, district_pre$malaria_blood_prevalence, paired = T)
```
The paired t-test results show a statistically significant difference between Blood Smear and Rapid Test measurements (\(t = -3.8688\), \(p = 0.00057\), \(df = 29\)), indicating that the mean of Rapid Test is higher than that of Blood Smear by approximately -0.206. The 95% confidence interval \([-0.3149, -0.0971]\) supports this conclusion, as it does not include 0. This suggests that Blood Smear consistently produces lower values than Rapid Test.

```{r}
aov(data = district_pre, malaria_rapid_prevalence ~ malaria_blood_prevalence)

```
The ANOVA results indicate that the predictor variable has a significant effect on the outcomes (\(F = 9.593\), \(p = 0.00301\)). The degrees of freedom for the test and residuals are 1 and 58, respectively, with the predictor explaining a variability of 0.976 (Sum Sq) compared to 5.900 for the residuals. The mean square for the predictor (\(0.9760\)) is notably higher than for the residuals (\(0.1017\)), and the highly significant p-value confirms that the predictor meaningfully influences the results. Together with the paired t-test, these findings show that the choice of test method significantly affects outcomes, with the paired t-test focusing on mean differences and ANOVA examining the overall variability.


### Bonus

2. Using a statistical model of your choice, determine if there is a relationship between malaria prevalence in a district and its average altitude. 
```{r}
# your code
model <- lm(BT_prev ~ `Average District Altitude`, data = dd)

# View the summary of the model
summary(model)



```


```{r}
library(ggplot2)

ggplot(dd, aes(x = `Average District Altitude`, y = BT_prev)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Relationship Between Malaria Prevalence and Altitude",
       x = "Average District Altitude (m)",
       y = "Malaria Prevalence") +
  theme_minimal()

```
There is a statistically significant relationship between district altitude and malaria prevalence. Higher altitudes are associated with lower malaria prevalence. However, the 
R-squared value indicates that altitude alone explains a small portion of the variability in malaria prevalence, suggesting other factors also play a significant role.
