---
title: "INTRODUCTION TO R PROGRAMMING"
author: "Rootkit"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Data Structures


```{r}
v<- c(1,0,0,T,T,F)
v
```
```{r}
x <- seq(0, 2*pi, length.out=90)
length(x)
range(x)
max(x)
min(x)
mean(x)
sum(x)
```

##looping with `rep` function
what are tail and head
```{r}
y <- rep(0:100)
head(y, 10)
tail(y, 10)

tail <- tail(y, 5)
```
```{r}
edit(tail)

```
```{r}
rep(c(0:1),each=10)
rep(c(0:1), c(10,10))
```
##Probabilities
Random generation

```{r}
set.seed(123)
sample(0:1, size=200, replace=TRUE, prob = c(1/3, 1-1/3)) -> y
sum(y ==0)
```
using `table()`
```{r}
t<- table(y)
t
```
proportions: `prop.table()` is applied on `table()`
```{r}
prop.table(t)
```
to get count we use `sum` and for the proportion we use `mean`
```{r}
sum(y == 0)
mean(y == 0)
```
In summary, `table()` can give us the count of all element in table format and `prop.table()` gives us proportions for each element and it have to be applied on `table()`

### Mimicking the LUDO game
```{r}
sample(6,size=1)
```
##Matrix
```{r}
matrix(rep(0, 5*6), nrow=5,)
```
1 0 0 1
0 1 1 1
0 0 1 1

$$
\begin{pmatrix}
1 & 0 & 0 & 1\\
0 & 1 & 1 & 1\\
0 & 0 & 1 & 1\\
\end{pmatrix}
$$

```{r}
elements <- c(1, 0, 0, 1,
              0, 1, 1, 1,
              0, 0, 1, 1)

# Create a 3x4 matrix
my_matrix <- matrix(elements, nrow = 3, ncol = 4, byrow = TRUE)

# Print the matrix
print(my_matrix)
```



