---
title: "dealing with outliers"
author: "Rootkit"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

lets generate dump data
```{r}
set.seed(42)
data <- sample(2.25:100, , replace = FALSE)
data[c(20,35)] <- data[c(20,25)] *10
data

boxplot(data)
data_copy <- data
```
```{r}
quat <- as.numeric(quantile(data, probs = c(0.25, 0.5, 0.75)))

df <- as.data.frame(t(quat))

names(df) <- paste0('Q', 1:length(quat))
rownames(df) <- "stats"
df$IQR <- df$Q3 - df$Q1
df$LowerLim = df$Q1 - 1.5*df$IQR
df$UpperLim = df$Q3 + 1.5*df$IQR
df
```
```{r}
(d_out <- data.frame(x = data, outlier = data < df$LowerLim | data > df$UpperLim))

```
identifying outliers
```{r}
data[d_out$outlier]
```
identifying outliers using boxplot
```{r}
(bxpt <- boxplot(data, range = 1.5, plot = FALSE)) # we do not plot
```
identifying outliers from the list
```{r}
# find all outliers
data[d_out$outlier] <- NA # or x[x %in% bxpt$out]
data[d_out$outlier] <- mean(x, na.rm = TRUE)

# boxplot of imputed x
par(mfrow = c(1, 2))
boxplot(data_copy, main = "Before\nhandling outliers", cex.main = 0.7)
boxplot(data, main = "After\nhandling outliers", col = "lightskyblue", cex.main = 0.7)
```
#Data simulation and visualization
## Chart in R
## Bar chart/plot
```{r}
fre <- c(Female=45, Children=20, Male=15)
barplot_values<-barplot(fre, col = rainbow(length(names(fre))), space = 0.001, legend.text = names(fre),ylim = c(0, max(fre) * 1.2) 
       )

text(x = barplot_values, y = fre, labels = fre, pos = 3, cex = 0.8, col = "black")

```

```{r}
par(mfrow = c(1, 2))
hist(iris$Sepal.Length, probability = T)
lines(density(iris$Sepal.Length), col="red", lwd=1.5)

plot(iris$Petal.Length,iris$Petal.Width, pch=16, col=iris$Species, probability = T, xlab = "Petal Length", ylab = "Petal Width")
#lines( lty = 2, lwd = 2, 0:7)
legend("topleft", legend = unique(iris$Species), pch = 16, col= unique(iris$Species), bty="n")

#let add a regression line to that fits the data
reg_model <- lm(Petal.Width ~ Petal.Length, data = iris)

# Add the regression line to the plot
abline(reg_model, col = "blue", lwd = 2, lty=2)


```
scatter plot to show the relationship btn two variables
```{r}
set.seed(42)
x<- runif(1000) # this provides 1000 random numbers

```

### Histogram
```{r}
hist(
  x, probability = T, main = "Histogram of a uniform distribution", col = 'lightskyblue'
  )
#adding a linr on histogram as an overlay
lines(density(x), lwd=2, lty=2)
```
calculating variance
```{r}
#1 type of variance
(meanx <- mean(x))
(variance1 <- var(x))
(variance2 <- mean(x^2) - meanx^2)
```
standard deviation
```{r}
(sdx <- sd(x))
#coefficient of 
(sdx/meanx)
```

Testing normarity
```{r}
ks.test(x,"pnorm")
qqnorm(x, col='black')
qqline(x, lwd=2, col='lightskyblue')
```
Binomial distribution, we can generate random numbers using `rbinom``
```{r}
rbinom_dist <- rbinom(10000, 1, 0.6)
hist(rbinom_dist, probability = TRUE, main = "Histogram of binomial distribution", 
     col = "turquoise", breaks = 20, xlab = NULL)
lines(density(rbinom_dist), col = "red", lwd = 2, lty = 2)
abline(v = mean(rbinom_dist), col = "blue", lty = 2, lwd = 2) # vertical line
legend("topleft", lty = c(2, 2), 
       col = c("red", "blue"), legend = c("Density", "Average"))
```

```{r}
par(mfrow=c(1,2))
pie(table(rbinom_dist))
barplot(table(rbinom_dist))
```
$$
f_x(x) = \frac{1}{\sqrt{2\pi\sigma}}\exp\left\{-\}
$$
