---
title: "assignment-1"
author: "Robert UWITONZE"
date: "2024-11-09"
output:   html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
data <- read.csv("../../data.csv")

```


```{r echo=FALSE}


data$Background[grepl("statistic", data$Background, ignore.case = TRUE)] <- "Applied Statistics"

data$Background[grepl("tourism", data$Background, ignore.case = TRUE)] <- "Travel and Tourism Management"

data$Background[grepl("data scie", data$Background, ignore.case = TRUE)] <- "Data Science"

data$Background[grepl("Information Technology Major of Information Management", data$Background, ignore.case = TRUE)] <- "Information Technology"

data$Background[grepl("Civil Engineering", data$Background, ignore.case = TRUE)] <- "Civil Engineering"

data$Background[grepl("Quantity sur", data$Background, ignore.case = TRUE)] <- "Quantity Surveying"

data$Background[grepl("design", data$Background, ignore.case = TRUE)] <- "Industrial Art and Design"

data$Background[grepl("information systems", data$Background, ignore.case = TRUE)] <- "Information Systems"



```

```{r}
 counts <- table(data$Background)
background_summary <- data.frame(
  Field = names(counts),
  Count = as.numeric(counts),
  Percentage = round(prop.table(counts) * 100, 2)
)
background_summary <- background_summary[order(-background_summary$Count), ]


background_summary$Percentage.Var1 <-NULL
colnames(background_summary) <- c("Field","Count", "Percentage")
background_summary
```


```{r}
?pie
```

```{r}
colors <- c("red", "orange", "yellow", "green", "cyan", "blue", "purple", "pink", "lightblue")

pie(
  background_summary$Percentage,
  labels =  paste(background_summary$Field, "(",background_summary$Percentage,"%)"),
  col = colors,
  cex=0.7
  )


```
<br>
Majority of participants are from **Applied Statistics** with <font color="red">*`52.94%`*</font>, 
**Data Science** follows with a substantial share of <font color="orange">*`9.8%`* </font>. 
 **Information Systems** and **Industrial Art and Design** each have a similar  percentage of participants of <font color="green">*`7.84%`*</font>. 
The remaining fields, including **Civil Engineering**, **Quantity Surveying**, **Travel and Tourism Management**, **Demography**, and **Information Technology**, have small percentages, ranging from <font color="purple">*`1.96%`*</font> to <font color = "blue">*`5.88%`*</font>. <br>
**Overall, the chart highlights a strong preference for fields related to data, statistics, and technology among the participants**.

