---
title: "Exercise 1"
author: "Lema Logamou Seknewna"
date: "`r Sys.Date()`"
output: pdf_document
---


Download the Rmarkdown [here](Exercises/Exercise1.Rmd).


1. Import the CSV (comma-separated value) using the `read.csv()` command. Name it `survey_data`.
```{r}
# write your code here
(survey_data = read.csv("data.csv"))

```


2. How many row and columns does `survey_data` contains.
```{r}
# write your code here
cat("the dataset has", nrow(survey_data), "rows and ",ncol(survey_data), "columns")

```


3. How many participants have a master's degree?
```{r}
# write your code here
cat(sum(survey_data["Completed.Program"] == "Master's degree"), "participants have master's degree")

```

4. How many participants have a bachelor's degree?
```{r}
# write your code here
cat(sum(survey_data["Completed.Program"] == "Bachelor's degree"), "participants have 	Bachelor's degree")
```

5. How many female participants have a bachelor's degree?
```{r}
# write your code here
cat(sum((survey_data["Completed.Program"] == "Bachelor's degree") & (survey_data["Gender"] == "Female")), " female participants have Bachelor's degree")
```


5. What is the percentage of male participants with a master's degree?
```{r}
# write your code here
( mastar_holder <- subset(survey_data, Completed.Program == "Master's degree"))
(master_holder_gender <- mastar_holder["Gender"])
(master_gender_table = table(master_holder_gender))
(master_prop_tabel = prop.table(master_gender_table))
cat("the percentage of Male participants with master's degree is ",master_prop_tabel["Male"]*100,"%")

```

6. Add a new column to calculate the age from date of birth
```{r}
# Assuming survey_data is your data frame and Date.of.Birth is in "day/month/year" format
(formaed_date <- as.Date(survey_data$Date.of.Birth, format = "%d/%m/%Y"))

# Calculate the age
(survey_data$Age <- as.numeric(difftime(Sys.Date(), formaed_date, units = "days")) / 365.25)
#survey_data

```


7. What is the percentage of male participants with under 30 years old?
```{r}
# write your code here
males <- subset(survey_data, Gender=="Male")
prop.table(table(males$Age < 30))[2] * 100

```

```{r}
# Filter rows where "information" is found in the Background column
(
  survey_data[grepl("UR", survey_data$Institution, ignore.case = TRUE) & 
                             grepl("Male", survey_data$Gender, ignore.case = TRUE), ]
)



```
## Outlier Detection
we user `IQR` to detect outliers and the formlar is $$[Q1 - 1.5*TQR, Q3 + 1.5*IQR]$$
```{r}


survey_data$Age[survey_data$Age < 10] <- NA
mean(survey_data$Age, na.rm = TRUE)

survey_data$Age[is.na(survey_data$Age)]  <- mean(survey_data$Age, na.rm = TRUE)
boxplot(survey_data$Age)
```


